Supplementary Information

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
update NIO results
@ValentineHerr , hopefully this won't create a merging conflict. I think the system can handle this now.

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 8)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs. `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (Feb 8)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T (CHTA)
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species with both
#LT:  2 species with both
#CB:  8 species with both
#SC:  1 species with both
```
Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 80% of site-species combinations (n=37 of 46; Fig. `r clim_sens`). 
There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), one for which only a temperature term was significant (*Chukrasia tabularis* at HKK), and none with no significant climate responses. 
Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 8)

###Variables selected: 
#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)

###other:
n_pos_first_order_P_term=42 #(exceptions: BEPA & TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=19/38*100
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 8)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb8)
n_DBHx_analyzed=204 #n rows in climate_interactions_coefficients, excluding Hansley
n_DBHx_analyzed_sig=88
#precip:
n_DBHxP_RW_analyzed= 36
n_DBHxP_RW_sig=16
n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season

n_DBHxP_BAI_analyzed= n_DBHxP_RW_analyzed
n_DBHxP_BAI_sig=17
n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-7 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
#T:
n_DBHxT_RW_analyzed= 32
n_DBHxT_RW_sig=12
n_DBHxT_RW_sig_neg=5 #notable that it's minority for RW, contrasting with slight majority neg for BAI

n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
n_DBHxT_BAI_sig=15
n_DBHxT_BAI_sig_neg=9 #mostly drought signal, including 4 species at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb8)
n_tested=37 #same for all metrics
n_best_model_RW=30
n_best_model_BAI=32
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
percent_neg_of_sig_trends="90-91%"
percent_sig_neg_trend_of_analyzed = "73-78%" 
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

>>>>

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs. `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (Feb 11)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T (CHTA)
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species just T 
#LT:  2 species with both
#CB:  8 species with both
#SC:  1 species with both
```
Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 78% of site-species combinations (n=36 of 46; Fig. `r clim_sens`). 
There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climate responses. 
Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 11)

###Variables selected: 
#PPT over PDF: 9 sites with both. PDF selected in HKK, HF, ZOF
#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)

###other:
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At LDW, the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`).

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 11)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb11)
n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
n_DBHx_analyzed_sig=90  #NIO water is included here
#precip:
n_DBHxP_RW_analyzed= 36
n_DBHxP_RW_sig=16
n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season

n_DBHxP_BAI_analyzed= n_DBHxP_RW_analyzed
n_DBHxP_BAI_sig=17
n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-6 #(exceptions: QURU & ACRU at HF, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
#T:
n_DBHxT_RW_analyzed= 32
n_DBHxT_RW_sig=12
n_DBHxT_RW_sig_neg=5 #notable that it's minority for RW, contrasting with slight majority neg for BAI

n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
n_DBHxT_BAI_sig=16
n_DBHxT_BAI_sig_neg=10 #mostly drought signal, including 4 species at CB
```

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb11)
n_tested=37 #same for all metrics
n_best_model_RW=31
n_best_model_BAI=32
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
percent_neg_of_sig_trends="90-91%"
percent_sig_neg_trend_of_analyzed = "73-78%" 
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
